How to write a PhD in Biological Sciences: a guide for the uninitiated

### 2021-07-20 2.16-foreign.Rmd
Add files via upload
# After copy edits

Reading this book you will learn of some biases in biological sciences and how you need to be aware of them in order to properly interpret your work. Be aware also that biological sciences is biased in where studies take place, and who publishes them [@culumber2019widespread; @nunez2021monolingual]. What we really need then are more studies from less studied parts of the world, and especially we need these to be researched and published by people from those countries. Why then do we insist that science be written in English?

English may be one of the most difficult languages to learn how to write as a foreign language because rules are so inconsistent, and the vocabulary is so large. At 171476 words, English has an enormous vocabulary, followed closely by Russian (150000), Spanish (93000) and Chinese (85568: although you should note that counting words isn’t very accurate and as for counting dictionary entries, Korean wins). I cannot defend the use of English as the language of science, but I can assert that having a universal language for communicating science is massively helpful, and makes a large swathe of the literature available to those scientists who are prepared to become proficient in a second language. However, none of this really helps you to prepare writing in English as a second language.

One of my 'pet hates' is that many non-English speaking editors seem to insist that you have your work checked by a 'native English speaker' (see Figure \@ref(fig:editor)). This suggests to me that some of the worst people to comment on whether or not something is written in good English, is another foreign language speaker. The lesson here is that you should be careful who you trust. Moreover, in my experience, having advice from a 'native English speaker' is no guarantee to getting a well written chapter or manuscript. Most of the world’s scientists did not grow up speaking English. Yet, rightly or wrongly, English is the language in which science is currently written. So, if English isn’t your mother tongue, should you expect to receive help during peer review?

@@

As I’ve stated above, I did not study English, and would be the first to admit that my grasp of grammar and syntax is by no means perfect (those of you with better English will have picked this up by now!). I have read enough correct English to know when something is incorrect, but that doesn’t mean that I know how to correct it. I have spent many long hours trying to decode what others have written, and in some cases this has involved me rewriting entire manuscripts. I still do this as a co-author, although I do remember asking one colleague to please send any further drafts in their native Spanish as it would be easier to translate than it would be to re-write.

The time component is at the crux of my reasoning why, as an editor, reviewer and advisor, I will not provide an English language service. It is both a time consuming and an unsatisfying experience. We all have our own voice, and correcting while maintaining other people’s voices is a painstaking task. There’s an entire profession that specialises in this (think translator). These days there are also services available from publishers to non-English authors to help them correct their English.

@@

Given that I’m not going to correct your English, what remarks would you expect me to make when reviewing a poorly written manuscript?

@@

Anyone you know who writes good English and is willing to help you. One option is the service offered by **AuthorAID** (http://www.authoraid.info/en/). Using AuthorAID, you can find a long term mentor who will help you with your English. You can read more about this approach here [@freeman2006publishing].

@@

Yes. It is essential that you use other languages to communicate your science with the wider community. Most of the world doesn't speak English, and so there is a real need for this to be done in your language. See a nice blog from the LSE on this subject here.

@@

>Kurt Vonnegut [-@vonnegut1982how]:
>"No matter what your first language, you should treasure it all your life. If it happens not to be standard English, and if it shows itself when you write standard English, the result is usually delightful, like a very pretty girl with one eye that is green and one that is blue."

>>>>

Reading this book, you will learn of some biases in biological sciences and how you need to be aware of them in order to properly interpret your work. Be aware also that biological sciences is biased in where studies take place, and who publishes them [@culumber2019widespread; @nunez2021monolingual]. What we really need then are more studies from less studied parts of the world, and we need these to be researched and published by people from those countries. Why then do we insist that science be written in English?

English may be one of the most difficult languages to learn to write as a foreign language because rules are so inconsistent, and the vocabulary is so large. At 171 476 words, English has an enormous vocabulary, followed closely by Russian (150 000), Spanish (93 000) and Chinese (85 568: although you should note that counting words isn’t very accurate and as for counting dictionary entries, Korean wins). I cannot defend the use of English as the language of science, but I can assert that having a universal language for communicating science is massively helpful, and makes a large swathe of the literature available to those scientists who are prepared to become proficient in a second language. However, none of this really helps you to prepare writing in English as a second language.

One of my 'pet hates' is that many non-English speaking editors seem to insist that you have your work checked by a 'native English speaker' (see Figure \@ref(fig:editor)). In my experience, having advice from a 'native English speaker' is no guarantee to getting a well-written chapter or manuscript. Most of the world’s scientists did not grow up speaking English. Yet, rightly or wrongly, English is the language in which science is currently written. So, if English isn’t your mother tongue, should you expect to receive help during peer review?

@@

As I’ve stated above, I did not study English, and would be the first to admit that my grasp of grammar and syntax is by no means perfect (those of you with better English will have picked this up by now!). I have read enough correct English to know when something is incorrect, but that doesn’t mean that I know how to correct it. I have spent many long hours trying to decode what others have written, and in some cases, this has involved me rewriting entire manuscripts. I still do this as a co-author, although I do remember asking one colleague to please send any further drafts in their native Spanish as it was easier to translate than it to re-write.

The time component is at the crux of my reasoning why, as an editor, reviewer and advisor, I will not provide an English language service. It is time consuming and an unsatisfying experience. We all have our own voice, and correcting while maintaining other people’s voices is a painstaking task. There’s an entire profession that specialises in this (think translator). These days there are also services available from publishers to non-English authors to help them correct their English.

@@

@@

Anyone you know who writes good English and is willing to help you. One option is the service offered by **AuthorAID** (http://www.authoraid.info/en/). Using AuthorAID, you can find a long-term mentor who will help you with your English. You can read more about this approach here [@freeman2006publishing].

@@

Yes. It is essential that you use other languages to communicate your science with the wider community. Most of the world doesn't speak English, and so there is a real need for this to be done in your language. Many journals now allow abstracts to be submitted bilingually. Obviously, the second language should have some relevance to the study itself. Even if there is nothing in the instructions to authors about submitting the abstract in another language, I'd encourage you to do it. The more we scientists are exposed to the reality that other languages exist, the more it will rub off on those that need to understand. [Read more on this blog @taskin2020long]

@@

>Kurt Vonnegut [-@vonnegut1982how]:
>
>"No matter what your first language, you should treasure it all your life. If it happens not to be standard English, and if it shows itself when you write standard English, the result is usually delightful, like a very pretty girl with one eye that is green and one that is blue."

### 2021-07-20 2.17-plagiarise.Rmd
Add files via upload
# After copy edits

## Making sure that you don’t plagiarise {#plagiarise}

Essentially plagiarism is when you copy somebody else's work. We're most commonly familiar with plagiarism of writing and these days this is especially easy with the copy and paste function. Most people are not even aware that they have plagiarised because sometime in the past they copied and pasted the work of somebody else into a document and this later became incorporated into their text without them being aware of it.

Plagiarism is a problem because essentially you are taking somebody else's work without attributing it to them.

### How to know if you have plagiarised

Today there are several pieces of software that are used to scan text that's written and available on the internet to discover plagiarism. One such example is Turnitin. One of the outputs of TurnItIn is to highlight text that matches other text already on the internet. As almost all journals published on the internet TurnItIn can accurately determine if text has been copied from another article or website.
I usually set TurnItIn to determine plagiarism with five or more consecutive words.

@@

###  What to do if plagiarism is detected in your work

@@

__Here's what you do:__
- Read the sentence that has been plagiarised several times to yourself. 
- Now without looking at that sentence, write another sentence that has the same meaning. Because it's very hard to replicate somebody else's words without copying them, what you should find is that you've written a fresh plagiarised sentence. 
- This can now be added to your text, changed as appropriately to fit your existing text. And that should be the end of your plagiarism worries.

__Here's what not to do:__
- Take the sentence swap out some of the words for synonyms and pass it off as your own. 
- The sentence will still have the same structure that you copied and essentially this is still somebody else's work. Moreover TurnItIn will still recognise this as plagiarism.

### How can you make sure that you never plagiarise?
Quite simply if you never cut and paste, you won’t plagiarise. It’s that easy. I understand why people copy and paste as a way to get started, or because someone else has written something so well, it’s hard to believe that you could ever write it any better. But actually, you can write it just as well, and writing it in your own words is worth so much more.

@@

It still seems amazing to me when I submit a student manuscript to TurnItIn and see that it is completely free of any plagiarism. There are so many words in English (>643 000 according to the Oxford English Dictionary) and so many different ways of putting them that you really can have your own writing style. Your writing style will be as unique to you as a fingerprint, and it will be entirely free of plagiarism. It's something you can celebrate.

Want to read more about plagiarism to understand what it is: Louw [-@louw2017defining]

### What is grey plagiarism?

I would maintain that there is not really any grey area in plagiarism. So what are people talking about when they refer to a grey zone? The nominal grey zone within plagiarism comes when you copy your own text, this is sometimes referred to as self-plagiarism, but this also refers to the associated practice of ‘manuscript recycling’ [@geraldi2021self-plagiarism]. Essentially this is really just the same as plagiarism but obviously if you are the original author then if you use your own text it's not the same as using somebody else's, because the work is still yours. However this is frequently complicated by multiple authors on a paper. For example, if I wrote a paper with five other people, even though the words were mine, if I use them again they would actually be attributed to all six authors and not just to me. Thus, if I use them again with five different authors it would still be plagiarism.

This is a particular problem in biological sciences (and some other sciences) when it comes to the methodology section ([see part 3](methods.html)). Many students believe after having written the [Materials and Methods](methods.html) once, they can’t possibly write it again in another way. Actually, you can and you should. It’s good practice, but remember that you can also cite the methods in a previous chapter. For example, chapter 2 has one method written out in great detail. Chapters 3 and 4 then reference chapter 2 while giving only a brief section on that methodology. You should already have seen some examples of how people do this in the literature, especially when citing their own previous work.

This is further complicated by having different publishers for your papers. Let's say I wrote a paper, I was the sole author and it was published by a journal that retained the copyright. If I use the same text again in another paper where again I was the soul author, but it was published in a different journal and they retained the copyright, that would be a breach of copyright to the first journal. It would be regarded as plagiarism.

Obviously, this situation really shouldn't occur because if you are the author of your own work then you should retain the copyright it should never be given to a publisher or a journal. The fact that many journals want to obtain copyright on work that they don't pay for is a complete disgrace but it's something that we're not going to go into right now (you can see rants here and here about science publishing scandals).

### Manuscript recycling

Manuscript recycling is the practice of re-publishing the same manuscript (or with minimal changes) in different journals. Manuscript recycling is a bigger problem in other subject areas, notably economics [@geraldi2021self-plagiarism], publishing issues in economics appear to need a lot more professionalism.

### Overall - just don’t plagiarise

>>>>

# Making sure that you don’t plagiarise {#plagiarise}

Plagiarism is when you copy somebody else's work. We're most familiar with plagiarism of writing and these days this is especially easy with the copy and paste function. Most people are not even aware that they have plagiarised because sometime in the past they copied and pasted the work of somebody else into a document and this later became incorporated into their text without them being aware of it.

Plagiarism is a problem because, essentially, you are taking somebody else's work without attributing it to them.

## How to know if you have plagiarised

Today there are several pieces of software that are used to scan text that's written and available on the internet to discover plagiarism. One such example is TurnItIn. One of the outputs of TurnItIn is to highlight text that matches other text already on the internet. As almost all journals publish on the internet, TurnItIn can accurately determine if text has been copied from another article or website. I usually set TurnItIn to determine plagiarism with five or more consecutive words.

@@

##  What to do if plagiarism is detected in your work

@@

__Here's what you do:__

- Read the sentence that has been plagiarised several times to yourself. 
- Now without looking at that sentence, write another sentence that has the same meaning. Because it's very hard to replicate somebody else's words without copying them, what you should find is that you've written a fresh unplagiarised sentence. 
- This can now be added to your text, changed as appropriate to fit your existing text. And that should be the end of your plagiarism worries.

__Here's what not to do:__

- Take the sentence swap out some of the words for synonyms and pass it off as your own. 
- The sentence will still have the same structure that you copied and, essentially, this is still somebody else's work. Moreover, TurnItIn will still recognise this as plagiarism.

## How can you make sure that you never plagiarise?
Quite simply, if you never cut and paste, you won’t plagiarise. It’s that easy. I understand why people copy and paste as a way to get started, or because someone else has written something so well, it’s hard to believe that you could ever write it any better. But actually, you can write it just as well, and writing it in your own words is worth so much more.

@@

It still seems amazing to me when I submit a student manuscript to TurnItIn and see that it is completely free of any plagiarism. There are so many words in English, and so many different ways of putting them that you really can have your own writing style. Your writing style will be as unique to you as a fingerprint, and it will be entirely free of plagiarism. It's something you can celebrate.

Want to read more about plagiarism to understand what it is: see Louw [-@louw2017defining]

## What is autoplagiarism?

I would maintain that there is not really any grey area in plagiarism. So what are people talking about when they refer to a grey zone? The nominal grey zone within plagiarism comes when you copy your own text, this is sometimes referred to as 'text-recycling', 'self-plagiarism' or 'autoplagiarism'. A new guideline from Hall et al. [-@hall2021understanding] sets out the different types of grey plagiarism, or text recycling:

### Developmental recycling {#develop}

Developmental recycling is when you are reusing text that you have written for example between your proposal and something you intend for publication or in an ethics application that you also want to use in your thesis. All of this sort of developmental recycling is permitted and actually encouraged. I would further encourage you to use the opportunity of recycling this text to develop it and refine it further, condensing and improving where you can.

### Generative recycling {#generative}

Generative recycling is where you take pieces of already published text for example from the methods when it does not make sense to change the text or actually makes it more obscure to reword it in order to avoid plagiarism. In my experience this doesn't amount to more than a few sentences describing technical settings on equipment. However this will depend strongly on your own subject area and may amount to larger chunks of text. I suggest that it is usually possible to reword most of the methods sections of papers. You should really only be generatively recycling material if you cannot avoid it: i.e. when the text becomes more obscure by your attempts to reword it.

### Adaptive recycling

Adaptive recycling is where you are using published text as the basis for a different type of content (e.g. a popular article online, a magazine, or op-ed). I think that this kind of text recycling is quite unnecessary because you almost certainly need to reword your text for a different audience. There maybe times such as figure legends where you need to reuse text that was already published. If you do find yourself in such a position then check with the copyright owner of the material that you are able to reuse the text that you want without legal issues.

### Duplicate recycling

Duplicate recycling (also known as 'manuscript recycling') is where large tracts of texts are essentially the same for the same message and audience. This is never likely to be sanctioned as it suggests that you are attempting to publish the same work twice. It will not be legal or ethical. Manuscript recycling is a bigger problem in other subject areas, notably economics [@geraldi2021self-plagiarism], publishing issues in economics appear to need a lot more professionalism.

## Overall - just don’t plagiarise

### 2021-07-20 2.18-phrasebank.Rmd
Add files via upload
# After copy edits

## Academic phrasebank {#phrasebank}

@@

When reading papers, I often come across the same phrases time and again. They seem so well honed to the situation that I’m envious about how the writer dreamt them up. Stealing them isn’t quite [plagiarism](plagiarism.html), but then they don’t regularly fit the exact situation that I’m looking for.

Imagine cutting and pasting them into a file, or even better a database that’s ready for any eventuality. Banking them over time so that they are always available.

@@

the academic phrasebank

### How to find the phrase that you need:

- Look across the top for different sections of text that you are currently writing.
- Then look down the left hand side to see the type of language function that you need. Click on something that looks applicable and watch the database open up.

### Getting started

@@

Is it plagiarism? Given that the chapter before is all about not copying and pasting, am I now telling you that it’s ok to copy and paste? Most of these phrases are less than five words or you will find that they are non-consecutive parts of a sentence. I don't think that most of these would constitute plagiarism if you pasted them into your document. However, I would suggest that you spend some time refining what you've written and that when you do that, some of these phrases may then come out.

>>>>

# Academic phrasebank {#phrasebank}

@@

When reading papers, I often come across the same phrases time and again. They seem so well honed to the situation that I’m envious about how the writer dreamt them up. Using them isn’t quite [plagiarism](plagiarism.html), but then they don’t regularly fit the exact situation that I’m looking for.

Imagine cutting and pasting these phrases into a file, or even better a database that’s ready for any eventuality. Banking them over time so that they are always available.

@@

[**the academic phrasebank**](http://www.phrasebank.manchester.ac.uk/)

## How to find the phrase that you need:

- Look across the top for different sections of text that you are currently writing.
- Then look down the left-hand side to see the type of language function that you need. Click on something that looks applicable and watch the database open up.

## Getting started

@@

Is it plagiarism? The preceding chapter is all about not copying and pasting, am I now telling you that it’s ok to copy and paste? Most of these phrases are less than five words or you will find that they are non-consecutive parts of a sentence. I don't think that most of these would constitute plagiarism if you pasted them into your document. However, I would suggest that you spend some time refining what you've written and that when you do that, some of these phrases may then come out.

### 2021-07-20 2.19-critical.Rmd
Add files via upload
# After copy edits

## Why critical reading is crucial for improving your writing {#critical}

It is hard to emphasise exactly how important your reading will be, when it comes to your writing. If you are sitting with a blank page in front of you and feel that you have nothing to draw on, then think again. All of the reading that you have done to get to this point has already helped you more than you think. As I’ve pointed out elsewhere, your chapter or manuscript is likely to resemble the formula that many others already follow, and standing on the shoulders of those giants will help you again. But it’s possible that your reading isn’t helpful if you aren’t being sufficiently ‘critical’ when you read: and by critical I am talking about the methodical approach that you take to reading that improves your comprehension of a paper (Figure \@ref(fig:Criticalreading)).

@@

### Making notes on good and bad styles

When you read, it’s worth making all sorts of comments about what you’re reading. It’s tempting to think that the only aspects of a paper that you should take notes on are the science (questions, approach and findings), but this is only part of your mandate as a critical reader. Of course you should be taking notes on the science, but you should also be using it as an opportunity to learn tips and tricks of writing. Once you are actively writing, and I’d suggest that this should have been right from the very beginning of your studies, you should be actively thinking about how the authors are writing. Are they writing well? Are parts written poorly? Then the critical part is to cogitate, albeit briefly: what works, what doesn’t and why.

@@

#### Some examples

@@

You don’t have to restrict yourself in what you can learn while reading other people’s papers. Think about the stages in writing a paragraph. Look at the paragraphs you are reading and see if you can spot the subject and summary sentences. Maybe they are absent from the entire paper, or maybe they are present in every paragraph. Does it help enhance the readability of the paper? What about the times when authors break the rules? Can you see in your reading that the majority of authors follow rules? Do they follow the formula? What is the power in shaking it all up? If it works, try and analyse why it works. Then you will be reading critically.

### Make handwritten notes

@@

### Recognising good or bad in what you read

It’s going to be hard to read and think at the same time. A little like walking and chewing gum, it comes with experience. And if you do fall over, stop chewing for a while and wait until you’ve got into your stride before having another go. Thinking is all part of the reading process, and while I agree that it might be hard to think on multiple levels while you read, it is possible. For example, consider what other worthy things you’ve thought about while reading this book. If you’ve not stopped to check your phone, or think about eating or drinking something, well done. It might be worth creating an atmosphere with less other stimulants while you read. If you only have your reading material and notebook, you’ll be able to concentrate all of your thinking on them instead of getting distracted.

@@

If you are looking for good fodder for your journal club, consider reviewing a preprint and publishing your review with a doi [@teixeiradasilva2017fortifying]. This will make the whole exercise a lot more worthwhile and produce a net output for your club.

>>>>

# Why critical reading is crucial for improving your writing {#critical}

It is hard to emphasise exactly how important your reading will be, when it comes to your writing. If you are sitting with a blank page in front of you and feel that you have nothing to draw on, then think again. All of the reading that you have done to get to this point has already helped you more than you think. As I’ve pointed out elsewhere, your chapter or manuscript is likely to resemble the formula that many others already follow, and standing on the shoulders of those giants will help you again. But it’s possible that your reading isn’t helpful if you aren’t being sufficiently ‘critical’ when you read: and by critical, I am talking about the methodical approach that you take to reading that improves your comprehension of a paper (Figure \@ref(fig:Criticalreading)).

@@

## Making notes on good and bad styles

When you read, it’s worth making all sorts of comments about what you’re reading. It’s tempting to think that the only aspects of a paper that you should take notes on are the science (questions, approach and findings), but this is only part of your mandate as a critical reader. Of course, you should be taking notes on the science, but you should also be using it as an opportunity to learn tips and tricks of writing. Once you are actively writing, and I’d suggest that this should have been right from the very beginning of your studies, you should be actively thinking about how the authors are writing. Are they writing well? Are parts written poorly? Then the critical part is to cogitate, albeit briefly: what works, what doesn’t and why.

@@

## Some examples

@@

You don’t have to restrict yourself in what you can learn while reading other people’s papers. Think about the stages in writing a paragraph. Look at the paragraphs you are reading and see if you can spot the subject and summary sentences. Maybe they are absent from the entire paper, or maybe they are present in every paragraph. Does it help enhance the readability of the paper? What about the times when authors break the rules? Can you see in your reading that the majority of authors the follow rules? Do they follow the formula? What is the power in shaking it all up? If it works, try and analyse why it works. Then you will be reading critically.

## Make handwritten notes

@@

## Recognising good or bad in what you read

It’s going to be hard to read and think at the same time. A little like walking and chewing gum, it comes with experience. And if you do fall over, stop chewing for a while and wait until you’ve got into your stride before having another go. Thinking is all part of the reading process, and while I agree that it might be hard to think on multiple levels while you read, it is possible. For example, consider what other worthy things you’ve thought about while reading this book. If you’ve not stopped to check your phone, or think about eating or drinking something, well done. It might be worth creating an atmosphere with less extraneous stimulants while you read. If you only have your reading material and notebook, you’ll be able to concentrate all of your thinking on them instead of getting distracted.

@@

If you are looking for good fodder for your journal club, consider reviewing a preprint and publishing your review with a [DOI](#DOI) [@teixeiradasilva2017fortifying]. This will make the whole exercise a lot more worthwhile and produce a net output for your club.

### 2021-07-20 2.20-proposal.Rmd
Add files via upload
# After copy edits

In my institution, you need to have your PhD or MSc approved by a group of 3 independent staff members (i.e. not including advisors) in your department. The department requires that your written proposal is publicly available to everyone in the department to look at (normally deposited in the secretary’s office) at least 7 days prior to the oral presentation of your proposal. This is not unique to my university, many institutions require that you formally propose the work that you plan to do in a PhD. They may all have very different requirements. It is well worth checking out what your institution requires and by when. Talk to your adviser about this.

Your proposal is a collection of documents, where each document is a chapter of the thesis, and is roughly equivalent to a published paper. Ideally, these should be bounded by an introduction, which gives the bigger picture of how the thesis is placed, and a time-line at the end that demonstrates how you can complete the thesis within the two or three years for your MSc or PhD, respectively. Note that your project proposal is placed on your permanent file within the department.

@@

We expect that your final thesis will be a set of papers (data chapters sandwiched between an [introduction and discussion](PhDintro.html)), and so we ask you to prepare the proposal in the same way. That is, you will normally have two (MSc) or five (PhD) data chapters. In the proposal, each chapter will consist of three parts:

@@

## Your Proposal - what is it good for?

@@

One PhD student estimated that she pasted 60-70% of her proposal text directly into her thesis. This included nearly all of the methods and materials section verbatim. Note that this **is not** plagiarism. You wrote the original proposal as a single author and it was not published, so you are at liberty to use the text again.

>>>>

In my institution, you need to have your PhD or MSc approved by a group of 3 independent staff members (i.e. not including advisors) in your department. The department requires that your written proposal is publicly available to everyone in the department to look at prior to the oral presentation of your proposal. This is not unique to my university, many institutions require that you formally propose the work that you plan to do in a PhD. They may all have very different requirements. It is well worth checking out what your institution requires and by when. Talk to your advisor about this.

Your proposal is a collection of documents, where each document is a chapter of the thesis. Ideally, these should be bounded by an introduction, which gives the bigger picture of how the thesis is placed, and a time-line at the end that demonstrates how you can complete the thesis within the two or three years for your MSc or PhD, respectively.

@@

Your final thesis will be a set of papers (data chapters sandwiched between an [introduction and discussion](PhDintro.html)), and so prepare the proposal in the same way. That is, you will normally have two (MSc) or five (PhD) data chapters. In the proposal, each chapter will consist of three parts.

@@

## Your proposal - what is it good for?

@@

One PhD student estimated that she pasted 60-70% of her proposal text directly into her thesis. This included nearly all of the methods and materials section verbatim. Note that this **is not** plagiarism, but [developmental recycling](#develop).
